---
output: github_document
bibliography:
  - "covariate_adjustment.bib"
---

<!-- impart.md is generated from README.Rmd. Please edit that file -->

```{r Setup, include = FALSE}
# Graphical Parameters
fig_w <- 8
fig_h <- 8

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.width = fig_w,
  fig.height = fig_h,
  fig.align = "center",
  out.width = "100%",
  dpi = 600,
  echo = FALSE,
  message = FALSE
)
```




## Installation

You can install the development version of impart from [GitHub](https://github.com/) using `install.packages` from the `devtools` package:

```{r Install Impart, echo = TRUE}
# install.packages("devtools") # Install `devtools` if it is not already installed
devtools::install_github("jbetz-jhu/impart")
```

## Vignettes

There are several vignettes on how to use `impart`, including group sequential designs and information adaptive designs. For group sequential designs, see the vignette below:

```{r, Vignettes-GSD, echo = TRUE, eval = FALSE}
vignette("Group_Sequential_Designs", package = "impart") # Design
```

### Information Monitored Designs

Each vignette is sequential, building upon the previous ones. Since the analysis approach depends on the type of outcome, there are vignettes for continuous, binary, and time-to-event outcomes. Each of which has example data built-in:

```{r Vignettes-Information-Adaptive, echo = TRUE, eval = FALSE}
vignette("Continuous_1_Study_Design", package = "impart") # Design
vignette("Continuous_2_Monitoring", package = "impart") # Information Monitoring
vignette("Continuous_3_Analyses", package = "impart") # Analyses

vignette("Binary_1_Study_Design", package = "impart") # Design
vignette("Binary_2_Monitoring", package = "impart") # Information Monitoring
vignette("Binary_3_Analyses", package = "impart") # Analyses

vignette("Time_To_Event_1_Study_Design", package = "impart") # Design
vignette("Time_To_Event_2_Monitoring", package = "impart") # Information Monitoring
vignette("Time_To_Event_3_Analyses", package = "impart") # Analyses
```




### Implementing New Methods with `impart`

Since new methods are constantly under development, `impart` can be extended by functions in other packages. A worked example of including new software is in the vignette below:

```{r, Vignettes-GSD, echo = TRUE, eval = FALSE}
vignette("Implementing_New_Methods", package = "impart") # Design
```
