---
title: "Monitored Analyses for a Continuous Outcome"
description: >
  Performing interim and final analyses for a continuous outcome
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Monitored Analyses for a Continuous Outcome}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r RMD-Setup, include = FALSE}
fig_w <- 8
fig_h <- 8

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  message = TRUE,
  error = TRUE,
  purl = FALSE,
  results = "markup",
  fig.path = "../man/figures/",
  fig.width = fig_w,
  fig.height = fig_h,
  fig.align = "center",
  out.width = "80%",
  dpi = 600
)

options(
  rmarkdown.html_vignette.check_title = FALSE
)

set.seed(9876)
```

```{r Library}
library(impart)
```

This vignette demonstrates how to use `impart` to conduct interim and final
analyses in group sequential or information monitored designs. This builds upon
the terminology and concepts in the vignettes on study design and information
monitoring. To see all available vignettes in impart, use the `vignettes` command:

```{r List-Vignettes, eval = FALSE}
print(vignette(package = "impart"))
```

## Setting Up a Monitored Design:

The design parameters in this vignette will be the same as in the previous
vignettes:

```{r Universal-Study-Design-Parameters}
# Universal Study Design Parameters
minimum_difference <- 5 # Effect Size: Difference in Means of 5 or greater
alpha <- 0.05 # Type I Error Rate
power <- 0.9 # Statistical Power
test_sides <- 2 # Direction of Alternatives

# Determine information required to achieve desired power at fixed error rate
information_single_stage <-
  impart::required_information_single_stage(
    delta = minimum_difference,
    alpha = alpha,
    power = power
  )

# Group Sequential Design Parameters
information_rates <-
  c(0.50, 0.75, 1.00) # Analyses at 50%, 75%, and 100% of the Total Information
type_of_design <- "asOF" # O'Brien-Fleming Alpha Spending
type_beta_spending <- "bsOF" # O'Brien-Fleming Beta Spending

# Set up group sequential testing procedure
trial_design <-
  rpact::getDesignGroupSequential(
    alpha = alpha,
    beta = 1 - power,
    sided = 2,
    informationRates = information_rates,
    typeOfDesign = type_of_design,
    typeBetaSpending = type_beta_spending,
    bindingFutility = FALSE
  )

# Inflate information level to account for multiple testing
information_adaptive <-
  impart::required_information_sequential(
    information_single_stage = information_single_stage,
    trial_design = trial_design
  )

# Initialize the monitored design
monitored_design <-
  initialize_monitored_design(
    trial_design = trial_design,
    null_value = 0,
    maximum_sample_size = 280,
    information_target = information_adaptive,
    orthogonalize = TRUE,
    rng_seed_analysis = 54321
  )
```


The data used in this example will also be the same: `impart::example_1`. Specifically, the data has been reverted to particular points in the study at which the information thresholds have been met:

  - `example_1_ia_1`: Data for Interim Analysis 1
  - `example_1_ia_2`: Data for Interim Analysis 2
  - `example_1_final`: Data for Final Analysis


```{r Load-Data-Interim-Analysis-1}
# Data for Interim Analysis 1
example_1_ia_1 <- impart::example_1_ia_1

head(example_1_ia_1)
```


## Interim Analysis 1

Before conducting the first interim analysis, the information level should be above the pre-specified threshold: the smoothed trajectory can be used to mitigate random variation in the information level.

Analysts will need to specify `estimation_function`, the function used to compute the estimate, and `estimation_arguments`, a list of arguments aside from the data needed for this computation. Estimators may have a variance correction factor that can be computed from the analysis parameters: this can be specified using the `correction_function` argument.

For the standardization estimator computed by `standardization()`, the corresponding small-sample variance correction is `standardization_correction()`: this adjusts the variance according to the sample size in each arm and the number of parameters in each regression model.


```{r Compute-Information-Trajectory}
# Obtain time of last event
last_event <-
  example_1_ia_1[, c(".e", ".t_1", ".t_2", ".t_3", ".t_4")] |>
  unlist() |>
  max(na.rm = TRUE) |>
  ceiling()

example_1_ia_1_prepared <-
  prepare_monitored_study_data(
    data = example_1_ia_1,
    study_time = last_event,
    id_variable = ".id",
    covariates_variables = c("x_1", "x_2", "x_3", "x_4"),
    enrollment_time_variable = ".e",
    treatment_variable = "tx",
    outcome_variables = c("y_1", "y_2", "y_3", "y_4"),
    outcome_time_variables = c(".t_1", ".t_2", ".t_3", ".t_4"), 
    # Observe missingness 1 week after target study visit
    observe_missing_times = c(30, 60, 90, 120) + 7
  )

data_ia_1_trajectory <- 
  information_trajectory(
    prepared_data = example_1_ia_1_prepared,
    monitored_design = monitored_design,
    estimation_function = standardization,
    estimation_arguments =
      list(
        estimand = "difference",
        y0_formula = y_4 ~ x_1 + x_2 + x_3 + x_4,
        y1_formula = y_4 ~ x_1 + x_2 + x_3 + x_4,
        family = gaussian,
        treatment_column = "tx",
        outcome_indicator_column = ".r_4"
      ),
    correction_function = standardization_correction,
    orthogonalize = TRUE,
    n_min = 50,
    n_increment = 3,
    rng_seed = 23456,
    control = monitored_analysis_control()
  )

data_ia_1_trajectory
```


Once the trajectory has been computed, it can be smoothed and plotted:


```{r Smooth-Information-Trajectory}
plot(
  information ~ y_4,
  data = data_ia_1_trajectory
)

abline(
  lm(
    formula = information ~ y_4,
    data = data_ia_1_trajectory
  ),
  lty = 1
)

# Requires `deming` package
abline(
  deming::theilsen(
    formula = information ~ y_4,
    data = data_ia_1_trajectory
  ),
  lty = 3
)

abline(
  h = monitored_design$original_design$information_thresholds,
  lty = 2
)
```


Once the information level has been confirmed to be above the threshold, conducting the analysis is similar to computing information: 


```{r Interim-Analysis-1}
interim_analysis_1 <-
  monitored_analysis(
    data = example_1_ia_1,
    monitored_design = monitored_design,
    estimation_function = standardization,
    estimation_arguments = 
      list(
        estimand = "difference",
        y0_formula = y_4 ~ x_1 + x_2 + x_3 + x_4,
        y1_formula = y_4 ~ x_1 + x_2 + x_3 + x_4,
        family = gaussian,
        treatment_column = "tx",
        outcome_indicator_column = ".r_4"
      ),
    correction_function = standardization_correction
  )

interim_analysis_1$interim_analysis_1$decision
interim_analysis_1$interim_analysis_1$decision_data
interim_analysis_1$interim_analysis_1$information_fraction_orthogonal
```




## Interim Analysis 2

All subsequent analyses are identical in syntax: a new dataset is provided, and
the result of the previous analysis is passed using the `monitored_design`
argument.


```{r Interim-Analysis-2}
example_1_ia_2 <- impart::example_1_ia_2

interim_analysis_2 <-
  monitored_analysis(
    data = example_1_ia_2,
    monitored_design = interim_analysis_1,
    estimation_function = standardization,
    estimation_arguments = 
      list(
        estimand = "difference",
        y0_formula = y_4 ~ x_1 + x_2 + x_3 + x_4,
        y1_formula = y_4 ~ x_1 + x_2 + x_3 + x_4,
        family = gaussian,
        treatment_column = "tx",
        outcome_indicator_column = ".r_4"
      ),
    correction_function = standardization_correction
  )

interim_analysis_2$interim_analysis_2$decision
interim_analysis_2$interim_analysis_2$decision_data
interim_analysis_2$interim_analysis_2$information_fraction_orthogonal
```







## Final Analysis

The syntax is identical for the final analysis:

```{r Final-Analysis}
example_1_final <- impart::example_1_final

final_analysis <-
  monitored_analysis(
    data = example_1_final,
    monitored_design = interim_analysis_2,
    estimation_function = standardization,
    estimation_arguments = 
      list(
        estimand = "difference",
        y0_formula = y_4 ~ x_1 + x_2 + x_3 + x_4,
        y1_formula = y_4 ~ x_1 + x_2 + x_3 + x_4,
        family = gaussian,
        treatment_column = "tx",
        outcome_indicator_column = ".r_4"
      ),
    correction_function = standardization_correction
  )

final_analysis$final_analysis$decision
final_analysis$final_analysis$decision_data
final_analysis$final_analysis$information_fraction_orthogonal
```

