---
title: "Monitored Analyses for a Binary Outcome"
description: >
  Performing interim and final analyses for a binary outcome
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Monitored Analyses for a Binary Outcome}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{rpact}
---

```{r RMD-Setup, include = FALSE}
fig_w <- 8
fig_h <- 8

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  message = TRUE,
  error = TRUE,
  purl = FALSE,
  results = "markup",
  fig.path = "../man/figures/",
  fig.width = fig_w,
  fig.height = fig_h,
  fig.align = "center",
  out.width = "80%",
  dpi = 600
)

options(
  rmarkdown.html_vignette.check_title = FALSE
)

set.seed(9876)
```

```{r Library}
library(impart)
```

This vignette demonstrates how to use `impart` to conduct interim and final
analyses in group sequential or information monitored designs. This builds upon
the terminology and concepts in the vignettes on study design and information
monitoring. To see all available vignettes in impart, use the `vignettes` command:

```{r List-Vignettes, echo = TRUE, eval = FALSE}
vignette(package = "impart")
```

```{r List-Vignettes-Table, echo = FALSE}
impart_vignettes <- vignette(package = "impart")
knitr::kable(impart_vignettes$results[, c("Title", "Item")])
```

**NOTE:** This section is under development and will be posted shortly.
