---
title: "Performing Monitored Analyses with `impart`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{impart_analyses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
fig_w <- 8
fig_h <- 8

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  message = TRUE,
  results = "markup",
  fig.path = "../man/figures/",
  fig.width = fig_w,
  fig.height = fig_h,
  fig.align = "center",
  out.width = "80%",
  dpi = 600
)

options(
  rmarkdown.html_vignette.check_title = FALSE
)
```

```{r setup}
library(impart)
```

This vignette demonstrates how to use `impart` to conduct interim and final
analyses in group sequential or information monitored designs. This builds upon
the terminology and concepts in the vignettes on study design and information
monitoring. These can be viewed as follows:

```{r Previous Vignettes, eval = FALSE}
vignette("impart_study_design", package = "impart") # Design
vignette("impart_monitoring", package = "impart") # Information Monitoring
```

## Setting Up a Monitored Design:

The design parameters in this vignette will be the same as in the previous
vignettes:

```{r Universal Study Design Parameters}
# Universal Study Design Parameters
minimum_difference <- 5 # Effect Size: Difference in Means of 5 or greater
alpha <- 0.05 # Type I Error Rate
power <- 0.9 # Statistical Power
test_sides <- 2 # Direction of Alternatives

# Determine information required to achieve desired power at fixed error rate
information_fixed <-
  impart::required_information_fixed_n(
    delta = minimum_difference,
    alpha = alpha,
    power = power
  )

# Group Sequential Design Parameters
information_rates <-
  c(0.50, 0.75, 1.00) # Analyses at 50%, 75%, and 100% of the Total Information
type_of_design <- "asOF" # O'Brien-Fleming Alpha Spending
type_beta_spending <- "bsOF" # O'Brien-Fleming Beta Spending

# Set up group sequential testing procedure
trial_design <-
  rpact::getDesignGroupSequential(
    alpha = alpha,
    beta = 1 - power,
    sided = 2,
    informationRates = information_rates,
    typeOfDesign = type_of_design,
    typeBetaSpending = type_beta_spending,
    bindingFutility = FALSE
  )

# Inflate information level to account for multiple testing
information_adaptive <-
  impart::required_information_sequential(
    information_fixed = information_fixed,
    trial_design = trial_design
  )

# Initialize the monitored design
monitored_design <-
  initialize_monitored_design(
    trial_design = trial_design,
    null_value = 0,
    maximum_sample_size = 280,
    information_target = information_adaptive,
    orthogonalize = TRUE,
    rng_seed_analysis = 54321
  )
```


The data used in this example will also be the same: `impart::example_1`

```{r}
head(impart::example_1)
```

## Interim Analysis 1

```{r Interim Analysis 1, eval = FALSE}
interim_analysis_1 <-
  monitored_analysis(
    data = data_interim_1,
    monitored_design = monitored_design,
    estimation_function = standardization,
    estimation_arguments = 
      list(
        estimand = "difference",
        y0_formula = y_4 ~ x_1 + x_2 + x_3 + x_4,
        y1_formula = y_4 ~ x_1 + x_2 + x_3 + x_4,
        family = gaussian,
        treatment_column = "tx"
      )
    )
```




## Interim Analysis 2

All subsequent analyses are identical in syntax: a new dataset is provided, and
the result of the previous analysis is passed using the `monitored_design`
argument.

```{r Interim Analysis 2, eval = FALSE}
interim_analysis_2 <-
  monitored_analysis(
    data = data_interim_2,
    monitored_design = interim_analysis_1,
    estimation_function = standardization,
    estimation_arguments = 
      list(
        estimand = "difference",
        y0_formula = y_4 ~ x_1 + x_2 + x_3 + x_4,
        y1_formula = y_4 ~ x_1 + x_2 + x_3 + x_4,
        family = gaussian,
        treatment_column = "tx"
      )
  )
```

