---
title: "Getting Started with `impart`"
description: >
  Learning about group sequential and information monitored designs with `impart`
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with `impart`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

There are several excellent resources available to provide both background on group sequential design methods and examples of planning and analyses using R:

  - The [`rpact`](https://www.rpact.org/) package for group sequential designs
    - [`rpact` vignettes](https://www.rpact.org/#vignettes)
  - The [`gsDesign`](https://keaven.github.io/gsDesign/) package for group sequential designs
    - [Group Sequential Designs Made Easy](https://keaven.github.io/gsd-shiny/) by Keaven Anderson and Nan Xiao

Since group sequential design methodology can also implemented in information monitoring trials, users should understand the fundamentals of these designs.




## Using `impart`

There are several vignettes built into `impart`: these include vignettes on information monitoring designs as well as group sequential designs. To see all available vignettes in impart, use the `vignettes` command:

```{r List-Vignettes}
print(vignette(package = "impart"))
```


If you are using `impart` in practice, please 

```{r Citing-Impart}
citation(package = "impart")
```

