---
title: "3. Performing Monitored Analyses with `impart`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{impart_analyses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r RMD-Setup, include = FALSE}
fig_w <- 8
fig_h <- 8

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  message = TRUE,
  error = TRUE,
  purl = FALSE,
  results = "markup",
  fig.path = "../man/figures/",
  fig.width = fig_w,
  fig.height = fig_h,
  fig.align = "center",
  out.width = "80%",
  dpi = 600
)

options(
  rmarkdown.html_vignette.check_title = FALSE
)

set.seed(9876)
```

```{r Library}
library(impart)
```

This vignette demonstrates how to use `impart` to conduct interim and final
analyses in group sequential or information monitored designs. This builds upon
the terminology and concepts in the vignettes on study design and information
monitoring. These can be viewed as follows:

```{r Previous-Vignettes, eval = FALSE}
vignette("impart_study_design", package = "impart") # Design
vignette("impart_monitoring", package = "impart") # Information Monitoring
```

## Setting Up a Monitored Design:

The design parameters in this vignette will be the same as in the previous
vignettes:

```{r Universal-Study-Design-Parameters}
# Universal Study Design Parameters
minimum_difference <- 5 # Effect Size: Difference in Means of 5 or greater
alpha <- 0.05 # Type I Error Rate
power <- 0.9 # Statistical Power
test_sides <- 2 # Direction of Alternatives

# Determine information required to achieve desired power at fixed error rate
information_single_stage <-
  impart::required_information_single_stage(
    delta = minimum_difference,
    alpha = alpha,
    power = power
  )

# Group Sequential Design Parameters
information_rates <-
  c(0.50, 0.75, 1.00) # Analyses at 50%, 75%, and 100% of the Total Information
type_of_design <- "asOF" # O'Brien-Fleming Alpha Spending
type_beta_spending <- "bsOF" # O'Brien-Fleming Beta Spending

# Set up group sequential testing procedure
trial_design <-
  rpact::getDesignGroupSequential(
    alpha = alpha,
    beta = 1 - power,
    sided = 2,
    informationRates = information_rates,
    typeOfDesign = type_of_design,
    typeBetaSpending = type_beta_spending,
    bindingFutility = FALSE
  )

# Inflate information level to account for multiple testing
information_adaptive <-
  impart::required_information_sequential(
    information_single_stage = information_single_stage,
    trial_design = trial_design
  )

# Initialize the monitored design
monitored_design <-
  initialize_monitored_design(
    trial_design = trial_design,
    null_value = 0,
    maximum_sample_size = 280,
    information_target = information_adaptive,
    orthogonalize = TRUE,
    rng_seed_analysis = 54321
  )
```


The data used in this example will also be the same: `impart::example_1`

```{r Example-1-Header}
head(impart::example_1)

# Obtain time of last event
last_event <-
  example_1[, c(".enrollment_time", ".t_1", ".t_2", ".t_3", ".t_4")] |>
  unlist() |>
  max(na.rm = TRUE) |>
  ceiling()

prepared_final_data <-
  prepare_monitored_study_data(
    data = example_1,
    study_time = last_event,
    id_variable = ".id",
    covariates_variables = c("x_1", "x_2", "x_3", "x_4"),
    enrollment_time_variable = ".enrollment_time",
    treatment_variable = "tx",
    outcome_variables = c("y_1", "y_2", "y_3", "y_4"),
    outcome_time_variables = c(".t_1", ".t_2", ".t_3", ".t_4"), 
    # Observe missingness 1 week after target study visit
    observe_missing_times = c(30, 60, 90, 120) + 7
  )
```

```{r Create-Interim-Analysis-Datasets}
data_n_interim_1 <-
  data_at_time_t(
    prepared_data = prepared_final_data,
    study_time = 783
  )

data_n_interim_2 <-
  data_at_time_t(
    prepared_data = prepared_final_data,
    study_time = 1318
  )

data_n_final <-
  data_at_time_t(
    prepared_data = prepared_final_data,
    study_time = 1478
  )

data_n_final <- data_n_final$data

# Subset to first 189 Outcomes
data_n_final <- data_n_final[which(rank(data_n_final$.t_4) <= 189),]
```




## Interim Analysis 1

```{r Interim-Analysis-1}
interim_analysis_1 <-
  monitored_analysis(
    data = data_n_interim_1$data,
    monitored_design = monitored_design,
    estimation_function = standardization,
    estimation_arguments = 
      list(
        estimand = "difference",
        y0_formula = y_4 ~ x_1 + x_2 + x_3 + x_4,
        y1_formula = y_4 ~ x_1 + x_2 + x_3 + x_4,
        family = gaussian,
        treatment_column = "tx",
        outcome_indicator_column = ".r_4"
      )
    )

interim_analysis_1$interim_analysis_1$decision
interim_analysis_1$interim_analysis_1$decision_data
interim_analysis_1$interim_analysis_1$information_fraction_orthogonal
```




## Interim Analysis 2

All subsequent analyses are identical in syntax: a new dataset is provided, and
the result of the previous analysis is passed using the `monitored_design`
argument.

```{r Interim-Analysis-2}
interim_analysis_2 <-
  monitored_analysis(
    data = data_n_interim_2$data,
    monitored_design = interim_analysis_1,
    estimation_function = standardization,
    estimation_arguments = 
      list(
        estimand = "difference",
        y0_formula = y_4 ~ x_1 + x_2 + x_3 + x_4,
        y1_formula = y_4 ~ x_1 + x_2 + x_3 + x_4,
        family = gaussian,
        treatment_column = "tx",
        outcome_indicator_column = ".r_4"
      )
    )

interim_analysis_2$interim_analysis_2$decision
interim_analysis_2$interim_analysis_2$decision_data
interim_analysis_2$interim_analysis_2$information_fraction_orthogonal
```







## Final Analysis

The syntax is identical for the final analysis:

```{r Final-Analysis}
final_analysis <-
  monitored_analysis(
    data = data_n_final,
    monitored_design = interim_analysis_2,
    estimation_function = standardization,
    estimation_arguments = 
      list(
        estimand = "difference",
        y0_formula = y_4 ~ x_1 + x_2 + x_3 + x_4,
        y1_formula = y_4 ~ x_1 + x_2 + x_3 + x_4,
        family = gaussian,
        treatment_column = "tx",
        outcome_indicator_column = ".r_4"
      )
    )

final_analysis$final_analysis$decision
final_analysis$final_analysis$decision_data
final_analysis$final_analysis$information_fraction_orthogonal
```

