---
title: "Implementing New Methods in `impart`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Implementing New Methods in `impart`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{pwr}
  %\VignetteDepends{ggplot2}
---

```{r RMD-Setup, include = FALSE}
fig_w <- 8
fig_h <- 8

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  message = TRUE,
  error = TRUE,
  purl = FALSE,
  results = "markup",
  fig.path = "../man/figures/",
  fig.width = fig_w,
  fig.height = fig_h,
  fig.align = "center",
  out.width = "80%",
  dpi = 600
)

options(
  rmarkdown.html_vignette.check_title = FALSE
)

set.seed(9876)
```


`impart` has several built-in methods for covariate adjustment, but is designed to work with methods from other R packages. This vignette will demonstrate how to integrate other software by writing a [wrapper function](https://en.wikipedia.org/wiki/Wrapper_function), and test the resulting function to make sure it works with `impart`.




## Notes for Developers

`impart` was written to minimize dependencies to avoid the potential problems that may arise when dependencies are deprecated or updated. This includes dependence on common, relatively mature packages such as the [tidyverse](https://www.tidyverse.org/) ecosystem.

While some code works identically for `tibble`s and base R `data.frame`s, these objects behave  differently under [vector subsetting](https://r4ds.had.co.nz/tibbles.html) - `my_tbl_df[, 1]` or `my_tbl_df[, 1:2]` return a `tbl_df` whereas `my_df[, 1]` returns a vector and `my_df[, 1:2]` returns a `data.frame`. Results of `dim`, `length`, `nrow`, `ncol`, and other functions may differ between a `tbl_df` and a `data.frame`: if this is not handled appropriately, this can result in explicit errors (resulting in a `stop`) or silently returning erroneous results. The connection between error messages and the cause of the error is often unclear without looking at source code.


For this reason, `impart::monitored_analysis` checks to ensure that data are passed as a `data.frame`: a `tbl_df` can be easily coerced to a `data.frame` using `as.data.frame()`. If some function depends on having a `tbl_df` and is unable to accept a `data.frame`, the wrapper should include a step to coerce the input back to a `tbl_df` before being passed to the function for analysis.




## Implementing `speff2trial::speffSurv()` in `impart`

The `speff2trial::speffSurv()` in the [`speff2trial` package on CRAN](https://cran.r-project.org/web/packages/speff2trial/index.html) computes a covariate-adjusted marginal [hazard ratio](https://covariateadjustment.github.io/estimands.html#survival_estimands). The [repo for this package is hosted on Github](https://github.com/mjuraska/speff2trial). This vignette shows how to create a wrapper for this function, which is already integrated in the package as `impart::speffsurv_impart()`. 

The `impart::colon_cancer` dataset will be used to test the wrapper, using the outcome of `years_to_death`: a time-to-event outcome that is subject to censoring. 

```{r Load-Colon-Cancer-Data}
library(impart)

# Extract two treatment arms: Lev+5FU (Chemotherapy) and Obs (Observation)
colon_cancer_5fu_vs_obs <-
  subset(
    x = colon_cancer,
    arm %in% c("Lev+5FU", "Obs")
  ) |>
  droplevels()

head(colon_cancer_5fu_vs_obs)
```




The wrapper function should take in all of the necessary arguments to address missing data, compute the estimate, and return the result. The result should either be a `numeric` vector named `estimate`, or a `list` containing a numeric vector named `estimate`. The value of `estimate` is extracted by `impart::calculate_estimate()` to obtain estimates and bootstrap confidence intervals.

First, run the function that requires implementation: see the documentation of the function (e.g. `?speff2trial::speffSurv`) for details:




```{r Run-Example}
library(speff2trial)

speffsurv_example <-
  speff2trial::speffSurv(
    formula =
      survival::Surv(time = years_to_death, event = event_death) ~
      age + sex + obstruction + perforation + organ_adherence + positive_nodes +
      differentiation +local_spread,
    data = colon_cancer_5fu_vs_obs,
    trt.id = "arm",
    fixed = TRUE
  )
```


The `summary` or `print` methods can be helpful in identifying the appropriate value to extract:


```{r Summarize-Example-Output}
summary(speffsurv_example)
```


Next, inspect the output: 


```{r Inspect-Output}
names(speffsurv_example)

str(speffsurv_example)
```


This function returns a `list`, and the marginal hazard ratio estimates are in a named numeric vector called `beta`.




## An Example Wrapper

A wrapper is needed to take in the appropriate arguments, compute the estimate, and then return a vector named `estimate` or a list containing that vector. 

```{r Example-Wrapper}
speffsurv_impart_demo <-
  function(
    data,
    estimand = "log_hazard_ratio",
    formula,
    treatment_column = NULL,
    alpha = 0.05,
    ci = FALSE
  ){
    if(!(estimand %in% c("log_hazard_ratio"))){
      stop("`estimand` must be \"log_hazard_ratio\".")
    }

    if(!(treatment_column %in% names(data))){
      stop("`treatment_column` (", treatment_column, ") must be in `data`.")
    }

    # Check for event indicators that are missing times, vice versa
    outcome_cols <-
      all.vars(update(old = formula, new = . ~ 0))
    
    if(length(outcome_cols == 2)){
      miss_rows <- which(rowSums(is.na(data[, outcome_cols])) == 1)
      if(length(miss_rows) > 0){
        stop("Indicators missing an event time or event times missing an",
             "outcome indicator: rows ", paste0(miss_rows, collapse = ", "))
      }
    }
    
    # Get baseline covariates from formula
    baseline_covariates <-
      all.vars(update(old = formula, new = 0 ~ .))

    # Impute any missing values using mean/mode imputation
    data <-
      impute_covariates_mean_mode(
        data = data,
        baseline_covariates = baseline_covariates
      )

    # Subset to individuals whose outcomes have been assessed:
    speffsurv_result <-
      speff2trial::speffSurv(
        formula = formula,
        data = data,
        trt.id = treatment_column,
        conf.level = 1 - alpha,
        fixed = TRUE
      )

    # Return CI if specified
    if(ci){
      speffsurv_summary <-
        data.frame(summary(speffsurv_result)$tab)
      lcl <- speffsurv_summary["Speff", "LowerCI"]
      ucl <- speffsurv_summary["Speff", "UpperCI"]
    } else {
      lcl = NULL
      ucl = NULL
    }

    return(
      list(
        estimate = as.numeric(speffsurv_result$beta["Speff"]),
        se = sqrt(as.numeric(speffsurv_result$varbeta["Speff"])),
        lcl = lcl,
        ucl = ucl,
        alpha = alpha,
        estimand = estimand
      )
    )
  }
```


Compare the results of the wrapper to the summary:


```{r Test-Wrapper}
summary(speffsurv_example)

speffsurv_impart_demo(
  data = colon_cancer_5fu_vs_obs,
  estimand = "log_hazard_ratio",
  formula = 
    survival::Surv(time = years_to_death, event = event_death) ~
    age + sex + obstruction + perforation + organ_adherence + positive_nodes +
    differentiation + local_spread,
  treatment_column = "arm",
  alpha = 0.05,
  ci = FALSE
)

speffsurv_impart_demo(
  data = colon_cancer_5fu_vs_obs,
  estimand = "log_hazard_ratio",
  formula = 
    survival::Surv(time = years_to_death, event = event_death) ~
    age + sex + obstruction + perforation + organ_adherence + positive_nodes +
    differentiation + local_spread,
  treatment_column = "arm",
  alpha = 0.05,
  ci = TRUE
)
```


The wrapper should also be tested on missing data, making sure they are addressed successfully:


```{r Test-Missing}
colon_cancer_5fu_vs_obs_mcar <- colon_cancer_5fu_vs_obs

# Set some categorical/continuous variables to NA
set.seed(seed = 12345)
miss_rows <- sample(x = 1:nrow(colon_cancer_5fu_vs_obs), size = 10)
colon_cancer_5fu_vs_obs_mcar$age[miss_rows] <- NA
colon_cancer_5fu_vs_obs_mcar$sex[miss_rows] <- NA
colon_cancer_5fu_vs_obs_mcar$positive_nodes[miss_rows] <- NA

speffsurv_impart_demo(
  data = colon_cancer_5fu_vs_obs_mcar,
  estimand = "log_hazard_ratio",
  formula = 
    survival::Surv(time = years_to_death, event = event_death) ~
    age + sex + obstruction + perforation + organ_adherence + positive_nodes +
    differentiation + local_spread,
  treatment_column = "arm",
  alpha = 0.05,
  ci = TRUE
)
```




## Testing the Wrapper

All wrappers should be appropriately tested: the [testthat package](https://testthat.r-lib.org/) provides a suite of functions that simplify specifying and executing a testing workflow.

```{r Load-testthat}
library(testthat)
```

A suggested test workflow includes, but is not limited to:

  - Testing for missing covariate data
  - Testing for missing outcome data
    - For time-to-event, checking both event indicators and event times
  - Checking for compatibility with input that contains:
    - `numeric`, `integer`, `logical`, `factor`, `string`
  - Checking for [`labelled` data from `haven`](https://haven.tidyverse.org/index.html)




### 1. Complete Data

```{r Test-1-Complete-Data}
test_that(
  desc = "Complete Data",
  code =
    {
      expect_no_condition(
        object =
          speffsurv_impart_demo(
            data = colon_cancer_5fu_vs_obs,
            estimand = "log_hazard_ratio",
            formula = 
              survival::Surv(time = years_to_death, event = event_death) ~
              age + sex + obstruction + perforation + organ_adherence + positive_nodes +
              differentiation + local_spread,
            treatment_column = "arm",
            alpha = 0.05,
            ci = FALSE
          )
      )
    }
)
```




### 2. Missing Covariates

```{r Test-2-Missing-Covariates}
test_that(
  desc = "Missing Covariates",
  code =
    {
      set.seed(seed = 12345)
      colon_cancer_5fu_vs_obs_mcar <- colon_cancer_5fu_vs_obs
      miss_rows <- sample(x = 1:nrow(colon_cancer_5fu_vs_obs), size = 10)
      colon_cancer_5fu_vs_obs_mcar$age[miss_rows] <- NA
      colon_cancer_5fu_vs_obs_mcar$sex[miss_rows] <- NA
      colon_cancer_5fu_vs_obs_mcar$positive_nodes[miss_rows] <- NA
      
      expect_no_condition(
        object =
          speffsurv_impart_demo(
            data = colon_cancer_5fu_vs_obs_mcar,
            estimand = "log_hazard_ratio",
            formula = 
              survival::Surv(time = years_to_death, event = event_death) ~
              age + sex + obstruction + perforation + organ_adherence + positive_nodes +
              differentiation + local_spread,
            treatment_column = "arm",
            alpha = 0.05,
            ci = TRUE
          )
      )
    }
)
```




### 3. Missing Outcome Indicator

```{r Test-3-Missing-Outcome-Indicator}
test_that(
  desc = "Missing Outcome Indicator",
  code =
    {
      set.seed(seed = 23456)
      colon_cancer_5fu_vs_obs_mcar <- colon_cancer_5fu_vs_obs
      miss_rows <- sample(x = 1:nrow(colon_cancer_5fu_vs_obs), size = 10)
      colon_cancer_5fu_vs_obs_mcar$event_death[miss_rows] <- NA

      expect_error(
        object =
          speffsurv_impart_demo(
            data = colon_cancer_5fu_vs_obs_mcar,
            estimand = "log_hazard_ratio",
            formula = 
              survival::Surv(time = years_to_death, event = event_death) ~
              age + sex + obstruction + perforation + organ_adherence + positive_nodes +
              differentiation + local_spread,
            treatment_column = "arm",
            alpha = 0.05,
            ci = TRUE
          ),
        regexp = "Indicators missing an event time"
      )
    }
)
```




### 4. Missing Outcome Times

```{r Test-4-Missing-Outcome-Times}
test_that(
  desc = "Missing Outcome Times",
  code =
    {
      set.seed(seed = 23456)
      colon_cancer_5fu_vs_obs_mcar <- colon_cancer_5fu_vs_obs
      miss_rows <- sample(x = 1:nrow(colon_cancer_5fu_vs_obs), size = 10)
      colon_cancer_5fu_vs_obs_mcar$years_to_death[miss_rows] <- NA

      expect_error(
        object =
          speffsurv_impart_demo(
            data = colon_cancer_5fu_vs_obs_mcar,
            estimand = "log_hazard_ratio",
            formula = 
              survival::Surv(time = years_to_death, event = event_death) ~
              age + sex + obstruction + perforation + organ_adherence + positive_nodes +
              differentiation + local_spread,
            treatment_column = "arm",
            alpha = 0.05,
            ci = TRUE
          ),
        regexp = "Indicators missing an event time"
      )
    }
)
```

